### 2022-08-24 article.tex
Progress

The line of code is simplified (as the result of the T test is never stored),
but the code is correct:
\verb|t.test| is indeed the name of an R function to do a T-test
and it is reasonable to assume that this code, when found in the code
of an article, is correct.

Here, however, we see a mismatch between the English description and the code:
by default, the R function \verb|t.test| does a \textbf{two}-tailed T-test.
A consequence of this fictional example is that the published p-values are
higher than needed, resulting in less significant findings, which
results in (needlessly) less conclusions drawn from a paper.

@@

One pseudorandom example of a mismatch between code and article
was found by me in \cite{labrecque2019interpretation}.
After the corresponding author of that paper was contacted,
the code was updated and confirmed to have the same
results as published in the paper.

@@

But regardless of the size of the code, 
it is the code that is the ground truth.

@@

There are multiple ways of publishing code,
that require increasingly more knowledge to do and result
in increasingly more reproducible experiments:
(1) publishing the code as text only, e.g. by printing
it in a Supplementary Materials, (2) publishing
the code on a hosted website, e.g on GitHub, with
added features (3) publishing 
the installed code within a virtual environment,
for example, using a Singularity container.
Here I will go into details of each of these ways.

@@

The raw code of a computational experiment is
more important to preserve than the article describing
the experiment, even when both describe the experiment identically,
correctly and reproducibly, 
as it has/should been that code that generated the results of an experiment.
Again: if the paper and code differ, it is the code that did
the actual work with the actual/true results.

@@

The simplest metric to assess the quality of code is 
the amount of (single) lines of code (SLOC). 
Code that has a lower SLOC count, i.e. it is more concise,
is usually regarded as having a lower chance of containing bugs.
Additionally, following a style guide (e.g. \cite{style_guide}), 
improves software quality \cite{fang2001}.

A more interesting metric is the cyclomatic complexity of the code.
The cyclomatic complexity is defined as the number of ways that
the code can be executed. 
For example, code that has only one \verb|if| statement
has a cyclomatic complexity of 2, as the condition within the \verb|if|
statement can be true or false.
The cyclomatic complexity correlates with code complexity,
where more complex code is likelier to contain or give rise to bugs 
\cite{abd2018calculating,chen2019empirical,zimmermann2008predicting}

A feature that deserves more attention
is that 'code must act as a teacher for future developers' \cite{sadowski2018modern}.
Error handling is one of the mechanisms to do so.
A simple example is when one (e.g. a BSc student) 
needs to calculate the variance of a distribution.
Commonly one needs to call a function to do so, 
e.g. \verb|var| in R, that needs the values of the distribution
as input.
However, to calculate the variance of a distribution, at least two values
are required. 
When the variance of (zero or) only one value is requested,
an error message can helpfully indicate that the function 
needs at least two values.

@@

Publishing raw code is already better than not publishing code at all,
as that raw code can be assumed to have run successfully at least once.
To reproduce an experiment from raw code can be painful, 
yet it is usually better than writing the same code from scratch.

@@

Code is sometimes published alongside a paper, 
in the form of a hosted website.
This way of publishing code gives additional options,
both for the authors of that code, as well as its users.
This practice has increased,
with the abstracts from articles in the academic journal
'Bioinformatics' referring to such a hosted website
has increased tenfold, going from below 5 percent in 2009 
to around 50 percent in 2017 \cite{russell2018large}.

@@

There are multiple websites for hosting code, with BitBucket, GitHub,
GitLab and SourceForge being the most well-known.
GitHub is currenly the most popular website to host code in general,
with millions of users.
GitHub users can create dedicated websites (called 'repositories')
to upload code. Code hosts, such as GitHub, 
store that uploaded code and provide ways 
for visiters to download the code and/or interact with it.
The use of such code hosting websites
accommodates collaboration \cite{perez2016ten}.
and improves transparency \cite{gorgolewski2016practical}.
An example from the author is the website \cite{bbbqarticleissue157},
that hosts part of the code for the paper \cite{bilderbeek2022transmembrane},
as shown in Figure \ref{fig:bbbqarticleissue157}.

@@

Hosted code has many additional features, such as
a version-controlled history, continuous integration scripts
and automatically generated metrics.
A code hosting website produces and shares this metadata.
This metadata can help to determine the
reproducibility, correctness and quality of code
of computational experiments.

@@

Hosted code improves transparency \cite{gorgolewski2016practical}.
The history of changes, called the commit history, 
helps to assess the honesty in the reporting of the code. 
Ideally, the full version history of the code 
has been generated, by hosting the code directly at its inception.
However, in some repositories, only the final version of the code has been
posted, which means all historical information is lost.
From a code history, it can be observed, for example, 
how much statistical tests have really/actually been done 
and see if all are reported:
failing to reporting a (usually non-significant) statistical tests is 
one of the many degrees of freedom 
in p-value hacking \cite{wicherts2016degrees}.

@@

Some code hosts, among others, GitHub, 
allows the researcher to start software scripts when uploading a new version.
To determine the reproducibility of the code, one such script should
run the code (on trivial data) to verify if the code works at all.
If the functionality of the code is lost, the researcher
can immediatly see which small change was the cause of this.
A formalized version of this workflow is called continuous integration (CI).
Using a CI service (such as GitHub Actions), is known to significantly 
increase the number of bugs exposed \cite{vasilescu2015} and increases
the speed at which new features are added \cite{vasilescu2015}.
Using CI opens up the possibility to informally annotate features of
a repository. For example, Figure \ref{fig:build_badge} shows a 
build badge that indicates that a LaTeX document (i.e. this paper)
could be built.

@@

Testing, in general, is an important mechanism to ensure
the correctness of code (an interesting example is \cite{rahman2020exploratory}
showing bugs in scientific software on the COVID-19 pandemic. 
It recommends more testing).
The percentage of (lines of) code tested is called the code coverage.
Code coverage correlates with code quality \cite{horgan1994,del1995correlation}. 
The non-profit organisation rOpenSci \cite{ram2018community},
which peer-reviews code,
has made it a prerequisite to have a code coverage of 100\%.
Also here, CI opens up the possibility to informally annotate the
code coverage. For example, Figure \ref{fig:badge_codecov} shows a 
build badge that indicates that certain R package's code 
has a 100\% code coverage.
Hence, code coverage is important metadata, that can automatically
be acquired when using code hosting website.

@@

Most code hosts, among others, GitHub and GitLab,
automatically keep track of certain metrics.
One such metric is the amount of code each contributor has made.
In that way, authors involved in the writing of the software
are honestly acknowledged. Figure \ref{fig:daisie_contributors}
shows the number of commits the authors of DAISIE \cite{etienne2020daisie}
have made to its code.
The distribution of contributed code per author 
is useful metadata, that can automatically
be acquired when using code hosting website.

@@

Albeit hosted code is vastly superior over raw code
in case of reproduciblity and correctness,
there are some problems with hosted code as well.
The most important one is that hosted code needs maintanance:
the libraries that are used by an experiment
will change over time, inevitably causing the experiment to fail
over time. 
The degradation of software is a known feature for nearly 
four decades (see the definition of 'bit rot' at \cite{steele1983hacker}),
here I will use the more modern (and in this context superior)
definition of software collapse \cite{hinsen2019dealing},
in which software fails due to dependencies on other 
software.
Where this paper recommends using older software foundations 
with multiple implementations \cite{hinsen2019dealing},
there is a modern solution that potentially has more longevity:
to publish the code installed in a virtual environment
(including the dependencies used at that time), as described below.

There is another problem that authors of academic code face,
which is that they will be contacted with a user sending in a bug report.
One solution for the author is to ignore such emails
and it can be argued that no energy should be wasted on published code
and work on something new instead.
However, upon the find of a new bug, the question is:
does all the research that use that software still 
result in the same conclusions?
A code hosting website usually has a place to report bugs
that offer transparency in the handling of bug reports.

>>>>

The line of code is simplified (as the result of the T test is never stored),
but the code is correct:
\verb|t.test| is indeed the name of an R function to do a T-test
and it is reasonable to assume that this code, when glancing over the code
of an article, is correct.

Here, however, we see a mismatch between the English description and the code:
by default, the R function \verb|t.test| does a \textbf{two}-tailed T-test.
A consequence of this fictional example is that the published p-values are
higher than needed, resulting in less significant findings, which
results in (needlessly) less conclusions drawn in this fictional paper.

@@

One pseudorandom example of a mismatch between code and article
was found by me in \cite{labrecque2019interpretation}.
After the corresponding author of that paper was contacted,
the code was updated and confirmed to have the same
results as published in the paper.
This sequence of events could only have happened, 
because the authors chose to publish their code.

@@

But regardless of the size of the code, 
it is the code that is the ground truth
and should be treated as a first-class citizen
in a computational experiment.

@@

There are multiple ways of publishing code,
that require increasingly more knowledge to do and result
in increasingly more reproducible experiments:
(1) publishing the code as text only, e.g. by printing
it in a Supplementary Materials, (2) publishing
the code on a hosted website, e.g on GitHub, with
added features (3) publishing 
the installed code within a virtual environment,
for example, using a Singularity container.
Here I will go into details of each of these ways,
by describing some advantages, disadvantages, as well as how to
make the code more FAIR data \cite{wilkinson2016fair}.

@@

The raw code of a computational experiment is
more important to preserve than the article describing
the experiment, even when both describe the experiment identically,
correctly and reproducibly, 
as it the code that can reproduce the results.

@@

The simplest metric to assess the quality of code is 
the amount of (single) lines of code (SLOC). 
Code that has a lower SLOC count, i.e. it is more concise,
is usually regarded as having a lower chance of containing bugs.
Additionally, following a style guide 
(e.g. the Tidyverse style guide \cite{wickham2019advanced} for R,
or PEP 8 \cite{van2001pep} for Python), 
improves software quality \cite{fang2001}.

A more interesting metric is the cyclomatic complexity of the code.
The cyclomatic complexity is approximately defined 
as the number of independent paths that
the code can be executed. 
For example, code that has only one \verb|if| statement
has a cyclomatic complexity of 2, as the condition within the \verb|if|
statement can be true or false,
resulting in the body of the \verb|if| statement being either
exectuted or ignored.
The cyclomatic complexity correlates with code complexity,
where more complex code is likelier to contain or give rise to bugs 
\cite{abd2018calculating,chen2019empirical,zimmermann2008predicting}

A feature that deserves more attention
is that 'code must act as a teacher for future developers' \cite{sadowski2018modern}.
Error handling is one of the mechanisms to do so.
A simple example is when one (e.g. a BSc student) 
needs to calculate the variance of a distribution.
Commonly one needs to call a function to do so, 
e.g. \verb|var| in R, that needs the values of the distribution
as input.
However, to calculate the variance of a distribution, at least two values
are required (else a division by zero occurs in the calcation of the variance). 
When the variance of (zero or) only one value is requested,
an error message can helpfully indicate that the function 
needs at least two values to work as expected.

@@

Publishing raw code is already better than not publishing code at all,
as that raw code can reasonably be assumed to have run successfully at least once.
To reproduce an experiment from raw code can be painful, 
yet it is usually better than writing the same code from scratch.
In general, it appears that sharing code is rare \cite{stodden2011trust,read2015sizing}.

% \paragraph{Raw code can be FAIR data}

It is hard to make raw code be FAIR data. It is hard for computers to find
copy-pasted code in a Supplementary Materials. 
That code is usually only accessible
if the paper has been published under an open access license.
When code can be found and accessed, it is an open question if the
code can \emph{still} interoperate with current datasets, software
and hardware. Raw code usually lacks the metadata to describe its needs.
I would say that raw code, albeit better than no code, fits
FAIR data only poorly.

@@

Code is sometimes published alongside a paper, 
in the form of a code hosting website.
This way of publishing code gives additional options,
both for the authors of that code, as well as its users.
This practice has increased,
with the abstracts from articles in the academic journal
'Bioinformatics' referring to such a hosted website
has increased tenfold, going from below 5 percent in 2009 
to around 50 percent in 2017 \cite{russell2018large}.

@@

There are multiple websites for hosting code, with BitBucket, GitHub,
GitLab and SourceForge being the most well-known.
GitHub is currenly the most popular website to host code in general,
with millions of users (a simple user search can confirm this).
GitHub users can create dedicated websites (called 'repositories')
to upload code. Code hosts, such as GitHub, 
store that uploaded code and provide ways 
for visiters to download the code and/or interact with it.
The use of such code hosting websites
accommodates collaboration \cite{perez2016ten}.
and improves transparency \cite{gorgolewski2016practical}.
An example from the author is the website \cite{bbbqarticleissue157},
that hosts part of the code for the paper \cite{bilderbeek2022transmembrane}
\iffalse
, as shown in Figure \ref{fig:bbbqarticleissue157}
\fi
.

@@

Hosted code has many additional features, such as
a version-controlled history, continuous integration scripts
and automatically generated metrics.
A code hosting website produces and shares this metadata.
This metadata can help to determine the
reproducibility, correctness and quality of code
of computational experiments.
Here, I discuss those features and why these are important.

@@

Hosted code improves transparency \cite{gorgolewski2016practical},
as, among others, the history of changes (i.e. the git commit history), 
helps to assess the honesty in the reporting of the code. 
Ideally, the full version history of the code 
has been generated, by hosting the code directly at its inception,
so that the ontology of the experiment can be viewed.
However, in some repositories, only the final version of the code has been
posted, which means all historical information is lost.
The code history has many important properties,
as, for example, 
it can be observed how much statistical tests have really/actually been done 
and see if all are reported:
failing to reporting a (usually non-significant) statistical tests is 
one of the many degrees of freedom 
in p-value hacking \cite{wicherts2016degrees}.

@@

Some code hosts, among others, GitHub, 
allow the researcher/user to start software scripts upon uploading a change.
This is useful to determine the reproducibility of the code: 
one such script should run the code (on trivial data) 
to verify if the code works at all.
If the functionality of the code is lost, the researcher
can immediatly see which small change was the cause of this.
A formalized version of this workflow is called continuous integration (CI).
Using a CI service (such as GitHub Actions for GitHub), 
is known to significantly 
increase the number of bugs exposed \cite{vasilescu2015} and increases
the speed at which new features are added \cite{vasilescu2015}.
Using CI opens up the possibility to informally annotate features of
a repository. For example, Figure \ref{fig:build_badge} shows a 
build badge that indicates that a \LaTeX document (i.e. this paper)
could be built.

@@

Testing, in general, is an important mechanism to ensure
the correctness of code (an interesting example is \cite{rahman2020exploratory}
showing bugs in scientific software on the COVID-19 pandemic:
it recommends more testing).
The percentage of (lines of) code tested is called the code coverage.
Code coverage correlates with code quality \cite{horgan1994,del1995correlation}. 
The non-profit organisation rOpenSci \cite{ram2018community},
which peer-reviews code,
has made it a prerequisite to have a code coverage of 100\%.
Also here, CI opens up the possibility to informally annotate the
code coverage. 
For example, build badges
\iffalse
(e.g. as shown in Figure \ref{fig:badge_codecov})
\fi
are visual indications on a website that can, among others,
show a certain R package's code has a 100\% code coverage.

Hence, code coverage is important metadata, that can automatically
be acquired when using code hosting website.

@@

Most code hosts, among others, GitHub and GitLab,
automatically keep track of certain metrics.
One such metric is the amount of code each contributor has made.
In that way, authors involved in the writing of the software
are honestly acknowledged. 
\iffalse
Figure \ref{fig:daisie_contributors}
shows the number of commits the authors of DAISIE \cite{etienne2020daisie}
have made to its code.
\fi
The distribution of contributed code per author 
is useful metadata, that can automatically
be acquired when using code hosting website.

@@

Albeit hosted code is vastly superior over raw code
in case of reproduciblity and correctness,
there are some problems with hosted code as well.
The most important one is that hosted code needs maintanance:
the libraries that are used by an experiment
will change over time, inevitably causing the experiment to fail
over time. 
The degradation of software is a known feature for nearly 
four decades (see the definition of 'bit rot' at \cite{steele1983hacker}),
here I will use the more modern (and in this context superior)
definition of software collapse \cite{hinsen2019dealing},
in which software fails due to dependencies on other 
software.
Where that paper recommends using older software foundations 
with multiple implementations \cite{hinsen2019dealing},
this paper recommends a modern solution that potentially has more longevity:
to publish the code installed in a virtual environment
(including the dependencies used at that time), as described below.

There is another problem that authors of academic code face,
which is that they may be contacted with a user sending in a bug report.
HIERO
One solution for the author is to ignore such emails
and it can be argued that no energy should be wasted on published code
and work on something new instead.
However, upon the find of a new bug, the question is:
does all the research that use that software still 
result in the same conclusions?
A code hosting website usually has a place to report bugs
that offer transparency in the handling of bug reports.
